Efficient Methods for Solving String Constraints with Large Repetition Bound

### 2023-03-13 2023-1-19/efficient_way_for_counting.tex
draft with concat

%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf]{acmart}
%% NOTE that a single column version is required for 
%% submission and peer review. This can be done by changing
%% the \doucmentclass[...]{acmart} in this template to 
%% \documentclass[manuscript,screen]{acmart}
%% 
%% To ensure 100% compatibility, please check the white list of
%% approved LaTeX packages to be used with the Master Article Template at
%% https://www.acm.org/publications/taps/whitelist-of-latex-packages 
%% before creating your document. The white list page provides 
%% information on how to submit additional LaTeX packages for 
%% review and adoption.
%% Fonts used in the template cannot be substituted; margin 
%% adjustments are not allowed.

\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary {arrows.meta, automata, positioning, fit}
\usepackage{url}
\usepackage{algorithm}
\usepackage{algpseudocode}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2022}
\acmYear{2022}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation emai}{June 03--05,
  2018}{Woodstock, NY}
%
%  Uncomment \acmBooktitle if th title of the proceedings is different
%  from ``Proceedings of ...''!
%
%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%  June 03--05, 2018, Woodstock, NY} 
\acmPrice{15.00}
\acmISBN{978-1-4503-XXXX-X/22/12}
% set algorithmicx

\renewcommand{\algorithmicrequire}{\textbf{Input:}}  % Use Input in the format of Algorithm  
\renewcommand{\algorithmicensure}{\textbf{Output:}} % Use Output in the format of Algorithm

@@

\newcommand*{\myvec}[1]{\bar{#1}}
\newcommand*{\regex}{\mathcal{R}}
\newcommand*{\strsort}{\verb|Str|}
\newcommand*{\intsort}{\verb|Int|}
\newcommand*{\lan}{\mathcal{L}}
\newcommand*{\highlight}[1]{\textbf{\textit{#1}}}
\newcommand*{\aut}{\mathcal{A}}
\newcommand*{\algfun}[1]{\texttt{#1}}
\newcommand*{\strlen}[1]{\texttt{len}(#1)}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

@@

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{DengHang Hu}
% \authornote{Both authors contributed equally to this research.}
\affiliation{%
  \institution{State Key Laboratory of Computer Science,
    Institute of Software, Chinese Academy of Sciences}
  \city{}
  \country{}
}
\email{hudh@ios.ac.cn}

\author{Zhilin Wu}
% \authornote{Both authors contributed equally to this research.}
\affiliation{%
  \institution{State Key Laboratory of Computer Science,
    Institute of Software, Chinese Academy of Sciences}
  \city{}
  \country{}
}
\email{wuzl@ios.ac.cn}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Regular expression with large repetition bounds frequently appears in the real world. However, state-of-art string solvers can not efficiently solve string constraints with large
  repetition bounds, which slows down string constraint
  applications like formal verification and symbolic execution. In this paper, we
  propose a new systematic algorithm based on the decision procedure in our previous work to solve the problem efficiently. We use
  heuristic ways like over-approximation and under-approximation to accelerate the search. Further, we extract amounts of regex expressions with repetition
  times from real-world programs and generate a significant benchmark. We evaluate the algorithm's performance on our generated
  and developed benchmarks. The result shows that our solver outperforms the state-of-art string solvers.

\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
% \begin{CCSXML}
% <ccs2012>
%    <concept>
%        <concept_id>10003752.10003766.10003776</concept_id>
%        <concept_desc>Theory of computation~Regular languages</concept_desc>
%        <ignificance>500</concept_significance>
%        </concept>
%  </ccs2012>
% \end{CCSXML}

\ccsdesc[500]{Theory of computation~Regular languages}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{string constraints, cost-enriched automata, regex expression, bounded repetition}

%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
% \begin{teaserfigure}
%   \includegraphics[width=\textwidth]{sampleteaser}
%   \caption{Seattle Mariners at Spring Training, 2010.}
%   \Description{Enjoying the baseball game from the third-base
%   seats. Ichiro Suzuki preparing to bat.}
%   \label{fig:teaser}
% \end{teaserfigure}

% \received{20 February 2007}
% \received[revised]{12 March 2009}
% \received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
Regular expressions are widely used in programming languages such as Javascript and Python. About 30–40 \% of Java, JavaScript, and Python software uses regex
matching\cite{redos_fse2019}. To find
potential software bugs automatically, many researchers prefer to use one of the
formal verification technology - symbolic execution \cite{symbolic_pldi2021}. Symbolic execution views symbolic paths as multi-theories
constraints and uses a constraints solver to solve these constraints. Thus the
efficiency and the ability of the constraints solver are vitally important.
However, the string theory is one of the most challenging theories in multi-theory constraints because it is easy to be undecidable\cite{undecidable_2016}. So the string constraints solver with the higher
ability (support more operators) usually performs slower. Most solvers try very hard to improve efficiency. For example, Z3 and cvc5 append many heuristic
derivation rules to speed up their search in the DPLL framework. Trau+ proposes
a new model called flat automaton and uses the CEGAR framework to under-approximate
and over-approximate the string constraints again and again.
Unfortunately, the string solvers above perform poorly when the string constraints contain large repetition bound. The DPLL-framework string solvers prefer to
search model strings with smaller lengths, but large repetition bound most often stands
for considerable model string lengths. The automaton-based solvers
unwind the automaton naively, so the search space of the unwound automaton is
exponential of the repetition bound. To address this issue, we extend
cost-enriched finite automaton\cite{atva2020} whose search space is linear to the
repetition bound. Our main contributions are:
\begin{enumerate}
  \item We extend cost-enriched finite automaton to model the extended regular expression with large repetition bound.
  \item We devise an efficient algorithm to solve string constraints with large repetition bound with the new automaton model.
  \item We generate many instances with large repetition bound from real-word regular expressions.
  \item We implemented our efficient algorithm on ostrich and compared it with state-of-art string solvers. The result shows the superiority of our automaton model and algorithm.
\end{enumerate}
%TODO: list the main contributions
The structure of this paper is as follows: We introduce some abbreviations and concepts in section \ref{sec:pre} and introduce the syntax and semantics of the string theory logic in section \ref{sec:str_t}. Then we develop cost-enriched finite automaton model and its operations at section \ref{sec:cefa}. The efficient algorithm is discussed in section \ref{sec:solve}, and the implementation of the algorithm is mentioned in section \ref{sec:imp}. Finally, we conclude our work and look into future work in section \ref{sec:conclu}.

\section{Preliminaries} \label{sec:pre}
%TODO: preliminaries
\highlight{Tokens.} We consider a finite \emph{alphabet} $\Sigma$, the set of all \emph{letters}. A
\emph{string} is a finite sequence of letters from $\Sigma$. We use $\Sigma^*$ to
denote the set of strings over $\Sigma$, $\epsilon$ to denote the empty string, and $a,
  b,\cdots$ to denote constant letters in $\Sigma$. We use $u, v,\cdots$ to denote constant
string and $x, y,\cdots$ to denote variable string. Moreover, we also consider a set of natural integer numbers $\mathbb{N} = \{0,1,2,3,\cdots\}$ and a set of integer numbers $\mathbb{Z}= \{\cdots, -1, 0, 1, \cdots\}$. We use $m,n,\cdots$ to
denote constant integer, and $i,j\cdots$ to denote variable integer. For vector, we use $\bar{v}$ to denote the vector of constant integer, $0_n$ to denote the zero vector with length $n$, $|\bar{v}|$ to denote the length of $|\bar{v}|$, and $v(i)$ to denote the integer value of $\bar{v}$
at position $i$, and $\cdot$ to denote the concatenation of two vectors.\newline
\highlight{CNF, DNF, clause, cube.} We assume that the reader is familiar with first-order logic. A \emph{literal} is an atomic proposition or its negation. Briefly, A \emph{clause} is a disjunction of literals. The empty clause is $false$. A formula is in \emph{conjunctive normal norm} (CNF) if it is a conjunction of clauses. A \emph{cube} is a conjunction of a consistent set of literals; The empty cube is $true$. A formula is in \emph{disjunctive normal form} (DNF) if it is a disjunction of cubes.\newline
\highlight{Basic Regular Language.} A nondeterministic finite state automaton (NFA) is a tuple $\aut=(Q, \Sigma, \delta, I, F)$ where $Q$ is a finite set of states; $\Sigma$ is a finite alphabet; $\delta\in Q\times\Sigma\times Q$ is the transition relation; $I, F\subseteq Q$ are the set of initial states and finite states respectively. We write a transition $(q, a, q') \in \delta$ as $q\xrightarrow{a} q'$ for readability. A \emph{run} of an NFA $\aut$ on a string $w = a_1\cdots a_n$ is a sequence of transitions $q_0\xrightarrow{a_1}q_1\cdots q_{n-1}\xrightarrow{a_n}q_n$ with $q_0\in I$. The run is \emph{accepting} if $q_{n}\in F$. A word $w$ is \emph{accepted} by an NFA $\aut$ if there is an accepting run on $w$. The \emph{language} of $\aut$ is the set of all words accepted by $\aut$, denoted by $\lan(\aut)$. From automata theory\cite{aut_hopcraft}, we know that the languages generated by basic regular expression are also in the class of $\lan(\aut)$. We call them \emph{Basic Regular Language}.
\section{String Theory with Extended Regex} \label{sec:str_t}
% TODO: Syntax and semantic
\subsection{Syntax}
In this paper, we extend string theory in our previous work \cite{atva2020} with extended regex. We call
the new theory as \textit{Extended String Logic (ESL)}. ESL contains the following
sorts: string sort \verb|Str|, integer sort \verb|Int|, the enumerable set of string
sort $\verb|Str|^m$, and the enumerable set of integer sort $\verb|Int|^m$ for each $m\in
  \mathbb{N}$. Furthermore, ESL includes the following predicate and function constants:
string concatenation $\verb|con|:\verb|Str|\times\verb|Str|\rightarrow \verb|Str|$;
string length $\verb|len|: \verb|Str|\rightarrow \verb|Int|$; regular membership $\in:
\verb|Str|\times \verb|Str|^m$; the basic regex operators such as Kleene, concatenation,  conjunction, and disjunction; the extended regex operator repetition and complement; and the usual constants of linear arithmetic.

The syntax of ESL is presented in Table\ref{tab:syntax}. $\varphi$ is a first-order logic formula that can be a regular membership $x\in\regex$,
a word equation $e$, a linear arithmetic (in)equality $\beta$, a negation of a formula, or a
disjunction of two formulas. A word equation $e$ is an equality of two string terms. The
(in)equality operator $\odot$ contains $=, \geq$. $\beta$ is an (in)equality
of two integer terms. A string term $s$ is an empty string $\epsilon$, letters
$a\in\Sigma$, string variable $x$, or a concatenation of two string terms. An integer
term $\alpha$ is an integer constant 0, integer constant 1, integer variable $i$, the
length of the string term $s$, minus an integer term, or plus of two integer terms. The
regular expression $\regex$ is built on empty string $\epsilon$, constant letter $a\in \Sigma$. The supported regex operations involve concatenation $\cdot$, disjunction $+$,
intersection $\times$, Kleene $*$, complement $C$, and repetition $\{m, n\}$.

We use function $FV(\varphi)$ to return all free variables of $\varphi$. We consider \emph{straight-line} fragment\cite{atva2020}\cite{quadratic_aplas2018} of this string theory logic. A word equation \emph{cube} $\bigwedge_{i=1}^n e_i$ is said to be straight-line if it can be rewritten as the form $\bigwedge_{j=1}^m x_j = s_j$ such that: (i) $x_1, \cdots, x_m$ are different variables; and (ii) $FV(s_j)\cap \{x_1, x_2,
  \cdots, x_{j-1}\} = \emptyset$. A formula $\varphi = (\bigwedge_{i=1}^n e_i) \wedge \Re \wedge P$
is said to be straight-line if $\bigwedge_{i=1}^n e_i$ is straight-line, $\Re$ is a
\emph{tube} of regular membership $x\in \regex$ or its negation $x\not\in \regex$, and $P$ is a Presburger arithmetic formula.

Note that we can convert an ESL formula to a DNF formula such that each term in the DNF
formula is the form $(\bigwedge_{i=1}^n e_i) \wedge \Re \wedge P$ where $P$ is a
Presburger formula. We call the form $(\bigwedge_{i=1}^n e_i) \wedge \Re \wedge P$ as \emph{normal form}. The straight-line fragment is similar to the SSA form of the
program.  Sometimes we use $s_1s_2$ to stand for $\verb|concat|(s_1,s_2)$ and $|x|$ to
stand for $\verb|len|(x)$.
\begin{table}
  \begin{tabular}{l}
    $\regex$ ::= $\epsilon\mid a\mid \regex\cdot\regex\mid \regex+\regex\mid \regex\times \regex\mid \regex^*\mid\regex^C \mid \regex\{m,n\}$ \\
    $\varphi$ ::= $x\in \regex\mid e\mid\beta\mid \neg\varphi\mid \varphi\vee\varphi$                                                      \\
    $e$ ::= $s_1=s_2 $                                                                                                                     \\
    $\beta$ ::= $\alpha_1\odot \alpha_2$                                                                                                   \\
    $s$ ::= $\emptyset \mid \epsilon \mid a\mid x\mid \verb|concat|(s_1, s_2)$                                                             \\
    $\alpha$ ::= $0\mid 1\mid i \mid  \verb|len|(s)\mid -\alpha\mid \alpha_1+\alpha_2$
  \end{tabular}
  \caption{Syntax of ESL}
  \label{tab:syntax}
\end{table}

\begin{example}[Straight-line Formula]
  \begin{equation} \label{eq:1}
    \varphi_1 ::= x = \verb|concat|(y,z) \wedge y = z \wedge x\in a\{1,100\}
  \end{equation}
  The word equation cube
  $c ::= (x = \verb|concat|(y,z) \wedge y = z)$ is straight-line because it santisfy : (i) $FV(z) = \{z\}$, (ii)$\{z\}\cap\{x\} = \emptyset$. So the formula $\varphi_1 = c\wedge  x\in a\{1,100\}$ is also straight-line.
\end{example}

\begin{example}[Non Straight-line Formula]
  \begin{equation} \label{eq:1}
    \varphi_2 ::=  x = \verb|concat|(y,z) \wedge y = \verb|concat|(x,z) \wedge x\in a\{1,100\}
  \end{equation}
  $\varphi_2$ is not straight-line because the word equation cube\newline
  $(x = \verb|concat|(y,z) \wedge y = z)$ is not straight-line.
\end{example}

\subsection{Semantics}
\highlight{Extended Regular Language} We have introduced basic regular language at
section \ref{sec:preliminaries}. In our paper, we add a new operator \emph{repetition} $\regex\{m,n\}$, which means repeating regex $\regex$ at least $m$ times and at most $n$ times. Repetition can be syntactically rewritten by concatenation and disjunction:
$\regex\{m,n\} \equiv \regex^m\mid\cdots\mid\regex^n$ where $\regex^k$ defines concatenate $k$ times($m\leq k\leq n$) to $\regex$. However, if we naively rewrite the repetition operation, the automaton size will be huge (linear to repetition times), causing the search space to be exponential. To address the issue, we proposed a new automaton model called \emph{Cost-Enriched Finite Automaton (CEFA)}, whose size is constant to repetition times. CEFA can not handle nested repetition ($\regex\{m,n\}$ or $\regex *$, where $\regex$ contains the repetition operator). So we have to rewrite it syntactically when the regex contains nested repetition (e.g., $a\{1,100\}\{2,2\}$ is rewritten to $a\{1,100\}a\{1,100\}$). Our regular expression is semantically equal to basic regular language but syntactically extended. We call the language of our regex \emph{Extended Regular Language} and denote it to $\lan^e(\regex)$.\newline
\highlight{Semantics} We assume that $S$ is the set of string variables over $\Sigma^*$, and $I$ is the set of integer variables. $\eta: S\times\Sigma\rightarrow\Sigma^*$ is the interpretation on string where $\eta(c)=c$ for every letter $c\in \Sigma$ and $\eta(s_1s_2)=\eta(s_1)\eta(s_2)$. $\theta: I\rightarrow\mathbb{Z}$ is the interpretation of arithmetic that is the same as that of Presburger arithmetic. Then the semantics is given by a satisfaction relation: $\eta, \theta\models \varphi$ defined in Table \ref{tab:semantics}. We say a formula $\varphi$ is \emph{satisfiable} if a solution exists $(\eta, \theta)$ such as $\eta, \theta\models \varphi$. A formula $\varphi$ is \emph{unsatisfiable} if no solution exists.

>>>>

\documentclass[runningheads]{llncs}

\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary {arrows.meta, automata, positioning, fit}
\usepackage{url}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[]{float}
% set algorithmicx
\renewcommand{\algorithmicrequire}{\textbf{Input:}}  % Use Input in the format of Algorithm  
\renewcommand{\algorithmicensure}{\textbf{Output:}} % Use Output in the format of Algorithm

@@

\newcommand*{\myvec}[1]{\vec{#1}}
\newcommand*{\regex}{\mathcal{R}}
\newcommand*{\strsort}{\verb|Str|}
\newcommand*{\intsort}{\verb|Int|}
\newcommand*{\lan}{\mathcal{L}}
\newcommand*{\highlight}[1]{\textbf{\textit{#1}}}
\newcommand*{\aut}{\mathcal{A}}
\newcommand*{\algfun}[1]{\texttt{#1}}
\newcommand*{\strlen}[1]{\texttt{len}(#1)}

@@

%%
\author{Denghang Hu\inst{1} \and
Zhilin Wu\inst{1}}
\authorrunning{D.Hu et al.}
% \authornote{Both authors contributed equally to this research.}

\institute{State Key Laboratory of Computer Science,
Institute of Software, Chinese Academy of Sciences, China\\
\email{\{hudh,wuzl\}@ios.ac.cn}}

\maketitle

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Regular expression with large repetition bounds frequently appears in the real world. However, state-of-art string solvers can not efficiently solve string constraints with such regular expression, which slows down string solver applications like formal verification and symbolic execution. In this paper, we
  propose a new systematic algorithm based on cost-enriched automaton. We use
  heuristic ways like over-approximation and under-approximation to accelerate the search. Further, we extract amounts of regex expressions with large repetition
  bounds from real-world programs and generate a significant benchmark. We evaluate the algorithm's performance on our developed benchmarks and other typical benchmarks. The result shows that our solver outperforms the state-of-art string solvers.

\keywords{string constraints, cost-enriched automata, regex expression, bounded repetition}
\end{abstract}

\section{Introduction}
Regular expressions are widely used in programming languages. About 30–40 \% of Java, JavaScript, and Python software uses regex matching \cite{redos_fse2019}. To handle the problems in software verification reasoning about string, the string theory is introduced and amounts of string solvers are developed.
However, the string theory is one of the most challenging theories because it is easy to be undecidable \cite{undecidable_2016}.
String solvers try very hard to improve efficiency:  DPLL(T)-based string solver Z3 \cite{z3} and Cvc5 \cite{cvc5} apply many heuristic derivation rules. Automaton-based string solver Trau+\cite{trau} proposed flat automaton and uses the CEGAR framework to under-approximate
and over-approximate the string constraints again and again.
Unfortunately, the string solvers above perform poorly when the string constraints contain large repetition bounds. [add a example]. To address this issue, we extend cost-enriched finite automaton(CEFA)\cite{atva2020} with linear arithmetical constraint. Use backward-propagating algorithm to search a solution. Our main contributions are as follows: 
\begin{enumerate}
  \item  Extend cost-enriched finite automaton to model the regular expression with large repetition bound.
  \item  Devise an efficient algorithm based on CEFA to solve string constraints with large repetition bound.
  \item  Generate many instances with large repetition bound from real-word regular expressions.
  \item  Implemented our efficient algorithm on Ostrich\ref{Ostrich} and compared it with state-of-art string solvers. The result shows the superiority of our automaton model and algorithm.
\end{enumerate}
%TODO: list the main contributions
% The structure of this paper is as follows: We introduce some abbreviations and concepts in section \ref{sec:pre} and introduce the syntax and semantics of the string theory logic in section \ref{sec:str_t}. Then we develop cost-enriched finite automaton model and its operations at section \ref{sec:cefa}. The efficient algorithm is discussed in section \ref{sec:solve}, and the implementation of the algorithm is mentioned in section \ref{sec:imp}. Finally, we conclude our work and look into future work in section \ref{sec:conclu}.

\section{Preliminaries} \label{sec:pre}
%TODO: preliminaries
\highlight{Tokens.} A finite \emph{alphabet} $\Sigma$ is the set of all \emph{letters}. A
\emph{string} is a finite sequence of letters from $\Sigma$. We use $\Sigma^*$ to
denote the set of strings over $\Sigma$, $\epsilon$ to denote the empty string, and $a,
  b,\cdots$ to denote constant letters in $\Sigma$. We use $u, v,\cdots$ to denote constant
string and $x, y,\cdots$ to denote variable string. Moreover, we also consider a set of natural integer numbers $\mathbb{N} = \{0,1,2,3,\cdots\}$ and a set of integer numbers $\mathbb{Z}= \{\cdots, -1, 0, 1, \cdots\}$. We use $m,n,\cdots$ to
denote integer constant, and $i,j\cdots$ to denote integer variable. For vector, we use $\myvec{v}$ to denote the vector of integer constant, $0_n$ to denote the zero vector with length $n$, $v(i)$ to denote the integer value of $\myvec{v}$ at position $i$, $\myvec{v_1}\cdot\myvec{v_2}$ to denote the concatenation of $\myvec{v_1}$ and $\myvec{v_2}$, $R$ to denote the vector of integer variable, $R_1 \cap R_2$ to denote the same variables of $R_1$ and $R_2$, and $|\myvec{v}|$ or $|R|$ to denote the length of vector.\newline
\highlight{CNF, DNF, clause, cube.} We assume that the reader is familiar with first-order logic. A \emph{literal} is an atomic proposition or its negation. A \emph{clause} is a disjunction of literals. The empty clause is $true$. A formula is in \emph{conjunctive normal norm} (CNF) if it is a conjunction of clauses. A \emph{cube} is a conjunction of a consistent set of literals; The empty cube is $false$. A formula is in \emph{disjunctive normal form} (DNF) if it is a disjunction of cubes.\newline
\highlight{Basic Regular Language.} A nondeterministic finite state automaton (NFA) is a tuple $\aut=(Q, \Sigma, \delta, I, F)$ where $Q$ is a finite set of states; $\Sigma$ is a finite alphabet; $\delta\in Q\times\Sigma\times Q$ is the transition relation; $I, F\subseteq Q$ are the set of initial states and finite states respectively. We write a transition $(q, a, q') \in \delta$ as $q\xrightarrow{a} q'$ for readability. A \emph{run} of an NFA $\aut$ on a string $w = a_1\cdots a_n$ is a sequence of transitions $q_0\xrightarrow{a_1}q_1\cdots q_{n-1}\xrightarrow{a_n}q_n$ with $q_0\in I$. The run is \emph{accepting} if $q_{n}\in F$. A word $w$ is \emph{accepted} by an NFA $\aut$ if there is an accepting run on $w$. The \emph{language} of $\aut$ is the set of all words accepted by $\aut$, denoted by $\lan(\aut)$. From automata theory\cite{aut_hopcraft}, basic regular expression only contains union, concatenation, and closure operations. We know that the language of basic regular expression are also in the class of $\lan(\aut)$. We call it \emph{Basic Regular Language}.
\section{String Theory with Regex} \label{sec:str_t}
% TODO: Syntax and semantic
In this section, the syntax and semantic of string theory is defined.
\subsection{Syntax}
In this paper, we propose a new logic called \textit{Extended String Logic (ESL)}. ESL contains the following
sorts: string sort \verb|Str|, integer sort \verb|Int|, the enumerable set of string
sort $\verb|Str|^m$, and the enumerable set of integer sort $\verb|Int|^m$ for each $m\in
  \mathbb{N}$. Furthermore, ESL includes the following predicate and function constants:
string concatenation $\verb|con|:\verb|Str|\times\verb|Str|\rightarrow \verb|Str|$;
string length $\verb|len|: \verb|Str|\rightarrow \verb|Int|$; regular membership $\in:
\verb|Str|\times \verb|Str|^m$; the regex operators such as closure, concatenation,  conjunction, disjunction, intersection, repetition and complement; and the usual constants of linear arithmetic.

The syntax of ESL is presented in Table \ref{tab:syntax}. $\varphi$ is a first-order logic formula that can be a regular membership $x\in\regex$,
a word equation $e$, a quantifier-free Presburger arithmetic formula $\beta$, a negation of a formula, or a
disjunction of two formulas. A word equation $e$ is an equality of two string terms.$\beta$ is an (in)equality of two integer terms and the (in)equality operator $\odot$ contains $=, \geq$. A string term $s$ is an empty string $\epsilon$, letters
$a\in\Sigma$, string variable $x$, or a concatenation of two string terms. An integer
term $\alpha$ is an integer constant 0, integer constant 1, integer variable $i$, the
length of the string term $s$, minus an integer term, or plus of two integer terms. The
regular expression $\regex$ is built on empty string $\epsilon$, constant letter $a\in \Sigma$. The supported regex operations involve concatenation $\cdot$, disjunction $+$,
intersection $\times$, closure $*$, complement $C$, and repetition $\{m, n\}$.

We consider \emph{straight-line} fragment\cite{atva2020}\cite{quadratic_aplas2018} $ESL_{s}$ of the string theory logic. 
A formula $\varphi$ is said to be straight-line if it can be converted to a DNF formula $\bigvee_{i} \phi_i$ and $\phi_i= (\bigwedge_{i=1}^n e_i) \wedge \Re \wedge P$
where $\bigwedge_{i=1}^n e_i$ is straight-line, $\Re$ is a
\emph{cube} of regular membership $x\in \regex$ or its negation $x\not\in \regex$, and $P$ is a Presburger arithmetic formula. A word equation \emph{cube} $\bigwedge_{i=1}^n e_i$ is said to be straight-line if it can be rewritten as the form $\bigwedge_{j=1}^m x_j = s_j$ such that:
\begin{enumerate}
  \item $x_1, \cdots, x_m$ are different variables,
  \item $FV(s_j)\cap \{x_1, x_2, \cdots, x_{j-1}\} = \emptyset$ where $FV(\varphi)$ are all free variables of $\varphi$. \label{straigth_line_condition_2}
\end{enumerate} 
We call straight-line formula $(\bigwedge_{i=1}^n e_i) \wedge \Re \wedge P$ \emph{normal form}.We use $s_1s_2$ to stand for $\verb|concat|(s_1,s_2)$, $|x|$ to
stand for $\verb|len|(x)$.
\begin{table}[H]
  \centering
  \begin{tabular}{l}
    $\regex$ ::= $\epsilon\mid a\mid \regex\cdot\regex\mid \regex+\regex\mid \regex\times \regex\mid \regex^*\mid\regex^C \mid \regex\{m,n\}$ \\
    $\varphi$ ::= $x\in \regex\mid e\mid\beta\mid \neg\varphi\mid \varphi\vee\varphi$                                                      \\
    $e$ ::= $s_1=s_2 $                                                                                                                     \\
    $\beta$ ::= $\alpha_1\odot \alpha_2$                                                                                                   \\
    $s$ ::= $\emptyset \mid \epsilon \mid a\mid x\mid \verb|concat|(s_1, s_2)$                                                             \\
    $\alpha$ ::= $0\mid 1\mid i \mid  \verb|len|(s)\mid -\alpha\mid \alpha_1+\alpha_2$ 
  \end{tabular}
  \caption{Syntax of ESL}
  \label{tab:syntax}
\end{table}

\begin{example}[Straight-line Formula]
  \begin{equation} \label{eq:1}
    \varphi_s ::= x =yz \wedge y = z \wedge x\in a\{1,100\} \wedge |x| = 1
  \end{equation}
  $\varphi_s$ is straight-line because we can split it to three parts: (i) a straight-line word equation cube $x =yz \wedge y = z$,
  (ii) a regular membership cube $x\in a\{1,100\}$, and (iii) a Presburger arithmetic formula $|x| = 1$.
\end{example}

\begin{example}[Non Straight-line Formula]
  \begin{equation} \label{eq:1}
    \varphi_{ns} ::=  x =yz \wedge y = xz \wedge x\in a\{1,100\} \wedge |x| = 1
  \end{equation}
  $\varphi_{ns}$ is not straight-line because the word equation cube\newline
  $x = yz \wedge y = xz$ does not satisfy condition \ref{straigth_line_condition_2}: $FV(yz) \cap \{y\} \not = \emptyset$ and $FV(xz) \cap \{x\} \not = \emptyset$.
\end{example}

\subsection{Semantics}
\highlight{Regular Language} We have introduced basic regular language at
section \ref{sec:pre}. In addition to basic regular operations, we syntactically support intersection, complement and repetition. As we all know, basic regular language is closed under intersection and complement. Furthermore, the operation \emph{repetition} $\regex\{m,n\}$ means repeating regex $\regex$ at least $m$ times and at most $n$ times. It can be syntactically rewritten by concatenation and union:
$\regex\{m,n\} \equiv \regex^m\mid\cdots\mid\regex^n$ where $\regex^k$ defines concatenate $k$ times($m\leq k\leq n$) to $\regex$. So the regular language defined in Table \ref{tab:syntax} is semantically equal to basic regular language. However, if we naively rewrite the repetition operation, the automaton size will linear to the repetition times, causing the search space to be exponential. To address the issue, we proposed a new automaton model called \emph{Cost-Enriched Finite Automaton (CEFA)}, whose size is not relevant to the repetition times. CEFA can not handle nested repetition (e.g., $\regex\{m_1,n_1\} \regex\{m_2,n_2\}$ or $(\regex\{m, n\}) *$) and complementation of repetition (e.g, $(\regex\{m,n\})^C$). So we have to rewrite these regexes syntactically (e.g., $a\{1,100\}\{2,2\}$ is rewritten to $a\{1,100\}a\{1,100\}$).\newline
\highlight{Semantics} We assume that $S$ is the set of string variables over $\Sigma^*$, and $I$ is the set of integer variables. $\eta: S\times\Sigma\rightarrow\Sigma^*$ is the interpretation on string where $\eta(c)=c$ for every letter $c\in \Sigma$ and $\eta(s_1s_2)=\eta(s_1)\eta(s_2)$. $\theta: I\rightarrow\mathbb{Z}$ is the interpretation of arithmetic that is the same as Presburger arithmetic. Then the semantics is given by a satisfaction relation: $\eta, \theta\models \varphi$ defined in Table \ref{tab:semantics}. We say a formula $\varphi$ is \emph{satisfiable} if a solution $(\eta, \theta)$ exists such as $\eta, \theta\models \varphi$. A formula $\varphi$ is \emph{unsatisfiable} if no solution exists.

\begin{table}[H]
  \centering
  \begin{tabular}{lcl}
     $\eta,\theta \models \varphi_1\vee \varphi_2$ & $\mathsf{iff}$ & $\eta,\theta \models \varphi_1 \text{ or } \eta,\theta \models \varphi_2$\\
     $\eta,\theta \models \neg\varphi $          & $\mathsf{iff}$ & $\eta,\theta \not\models \varphi$\\
     $\eta, \theta \models x\in \regex$            & $\mathsf{iff}$ & $\exists w \in \lan(\regex),\eta,\theta \models x = w$\\
     $\eta, \theta \models s_1 = s_2$              & $\mathsf{iff}$ & $\eta(s_1) = \eta(s_2)$\\
     $\eta, \theta \models \alpha_1\odot\alpha_2$  & $\mathsf{iff}$ & $\theta(\alpha_1) \odot \theta(\alpha_2) \text{ where } \odot = \{=, \geq \}$
  \end{tabular}
  \caption{Semantics}
  \label{tab:semantics}
\end{table}
\subsection{Problem}
In this paper, we consider the following problem:
\begin{table}[H]
  \centering
  \begin{tabular}{l l}
    \hline
    \textbf{Problem:} & Is an $ESL_s$ formula $\varphi$ satisfiable?                   \\
    \textbf{Input:}   & A straight-line $ESL_s$ formula $\varphi$ in the normal form. \\
    \textbf{Output:}  & \emph{sat} or \emph{unsat}.                               \\
    \hline
  \end{tabular}
\end{table}
